The \NAME Framework:\\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-10-17 paper.tex
started rewrite, added old id algo classification

\title{
Introducing the \NAME Framework\\
for Efficient Intrusion Detection\\
in the Internet of Things
}

@@

Along with its potential, the IoT also presents a significant threat: by
opening our smart homes \cite{aldrich2003smart} and personal data to all these
interconnected devices, we also open them to everyone who is able to break
these new, virtual locks, often without leaving a trace.

@@

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) steps in: an ID probe monitors all traffic that passes through
the firewall, looking for patterns and optionally alerting the firewall if a
malicious pattern is detected \cite{denning1987intrusion}.

@@

In mobile ad hoc networks (MANET), which make up an important part of the IoT,
it is not possible to have this single point in the network that can oversee
all traffic \cite{zhang2000intrusion, mishra2004intrusion}. If the outer
security perimeter diminishes, every device has to implement its own lines of
defense. Making the IDS a local service on each device requires resources,
which are another important constraint: IoT devices typically have limited
batteries, little processing power and memory is restricted to the bare minimum.

@@

The IoT is a heterogeneous environment, both in hardware and software.
Addressing this in a transparent and automated way is a key success factor for
any kind of development in this area. For a non-functional layer, such as ID,
this aspect becomes even more paramount.

A solution that supports the integration of multiple algorithms on IoT devices
needs to consider the impact on their resources. Different algorithms should be
merged to avoid accumulating their impact.

@@

The remainder of this paper proceeds as follows: section \ref{background}
details some of the ideas behind FCF and identifies the different aspects of
the problem space. Section \ref{design} describes the design we applied to
construct the framework, its domain specific language (DSL) and code generator.
Section \ref{evaluation} evaluates an implementation and application of the
framework. Section \ref{related} explores related work in the field of DSLs for
ID and code generation for the IoT. Finally, section \ref{conclusion}
summarizes our findings, draws conclusions and identifies topics for future
work.

\section{Background}
\label{background}

The underlying problem addressed in this paper is of an economic nature: adding
ID doesn't add functionality, so its impact should be reduced as much as
possible.

>>>>

\title{
The \NAME Framework:\\
Efficient Intrusion Detection\\
in the Internet of Things
}

@@

The IoT presents a great potential to positively influence our daily work and
life: domotics, assisted living, e-health and enhanced learning are only a few
examples\cite{atzori2010internet}. But with this potential the IoT also
presents a significant threat: by opening our smart homes
\cite{aldrich2003smart} and personal data\cite{weber2010internet} to all these
interconnected devices, we also open them to everyone who is able to break
these new, virtual locks. IoT devices typically have limited batteries, little
processing power and memory is restricted to the bare minimum. These properties
make it hard to add any additional service, including security.

@@

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) steps in: an ID probe monitors all traffic that passes through
the firewall, looking for patterns and optionally alerting the firewall if a
malicious pattern is detected\cite{denning1987intrusion}.

@@

In wireless networks of resource constraint devices, which make up an important
part of the IoT, it is not possible to have this single point in the network
that can oversee all traffic \cite{zhang2000intrusion, mishra2004intrusion}. If
the outer security perimeter diminishes, every device has to implement its own
lines of defense. Making the IDS a local service on each device requires
resources, which are already very constraint.

@@

The IoT is a heterogeneous environment, both in hardware and software.
Addressing this in a transparent and automated way is a key success factor for
any kind of development in this area. A solution that supports the integration
of multiple algorithms on IoT devices needs to consider the impact on their
resources. Different algorithms should be merged to avoid accumulating their
impact.

@@

The remainder of this paper proceeds as follows: section \ref{landscape} takes
a look at ID algorithms and classifies them. Section \ref{analysis} looks for
patterns in the implementation of these classes of algorithms, details some of
the ideas behind FCF and identifies the different aspects of the problem space.
Section \ref{design} describes the design we applied to construct the
framework, its domain specific language (DSL) and code generator. Section
\ref{evaluation} evaluates an implementation and application of the framework.
Section \ref{related} explores related work in the field of DSLs for ID and
code generation for the IoT. Finally, section \ref{conclusion} summarizes our
findings, draws conclusions and identifies topics for future work.

\section{Landscape of ID Algorithms}
\label{landscape}

Research into ID in wireless networks typically focuses on two major topics
that complement the architecture: the devices as single entities and the
network as a group of such devices. Both are important because the group cannot
make a decision without members that detect malicious behavior and a
group-based decision is often needed because nodes can easily miss out on
certain events that could indicate intrusions, due to their wireless and
not-always-on nature.

\subsection{Detecting Intrusions}
\label{subsection:detecting}

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: anomaly detection,
signature or misuse detection, and specification-based detection. In
\cite{alrajeh2013intrusion} the authors mostly agree with this topology, but
add the notion of hybrid intrusion detection systems and cross layer intrusion
detection systems as recent advances in research. Although these additional
categories offer very interesting prospects, the authors have to admit that the
impact on the resource-constrained sensor nodes might still be too high.

The meshed network topology of WSNs and its routing protocols have produced
many related attacks and each of these attacks has been the focus of many
research papers presenting algorithms to detect them. In the following
paragraphs we take a look at some popular ones.

In a wormhole attack, an artificial low-latency link is created between two
distant nodes in the network. This causes the routing protocol to divert more
traffic through this link, offering the attacker an abundance of packets to
inspect. In \cite{maheshwari2007detecting} an algorithm is proposed to allow
nodes to determine if a wormhole is actively present. In essence, the algorithm
relies on the exchange of neighbour lists to allow nodes to determine if
unexpected substructures are present in the graph representing the network.

Another attack that is related to the wormhole is the sinkhole. When launching
a sinkhole attack, as described in \cite{krontiris2008launching}, the attacker
first needs to compromise an existing node. Being in control of the node, the
attacker can make his captured node look more \emph{attractive} to the
surrounding nodes and thus draw more traffic to itself. The sinkhole attack is
typically a foundation for other routing level attacks, such as selective
forwarding, modification of packets or even dropping them altogether. In its
simplest form, the sinkhole offers the attacker with lots of packets to analyze
and use to gather information about the network and its functionality.

Detecting sinkholes is very hard and is often only possible based on other
attacks that are supported by the sinkhole. When the sinkhole is used to
implement selective forwarding, \cite{ngai2006intruder} proposes an algorithm
that allows a base station, the aggregating master node to which all nodes
typically send their information, to observe missing data from nodes in the
same area in a statistical way.

More typical attacks such as flooding, the sybil attack, rushing\dots are
amongst others presented in \cite{wood2002denial} and \cite{djenouri2005survey}.

\subsection{Cooperative Decision Making}
\label{subsection:coorperative}

As in many other situations, a group is often stronger than the sum of its
individual components. This is surely the case for WSNs. Because not all nodes
are always actively participating in the network, some nodes might miss clues
that would otherwise lead them to detect intrusions. It is hardly impossible
for a single node to detect a complex attack by itself. Therefore a second
important research topic consists of cooperative algorithms to combine
information from single nodes into a group-based decision about alleged
intrusions.

In \cite{krontiris2009cooperative}, the authors first present a theoretical
foundation to analyze cooperative algorithms. Given these foundations they
continue to present an algorithm consisting of 5 phases. It essentially
implements a voting system based on the Guy Fawkes protocol
\cite{anderson1998new} to identify an intruder. It enables the exchange of
suspected intruder information and allows distributed authentication of those
\emph{votes}. Based on these authenticated votes, a distributed decision can be
made about the commonly identified intruder.

A distributed, cooperative algorithm can sometimes be implemented locally. This
is illustrated by the reputation-based detection algorithm introduced in
\cite{ganeriwal2008reputation}. Using this algorithm, nodes can exchange
information about the reputation of other nodes. Combining this information
allows them to decide about their trust in a given node, based on more than
their own, often partial, observations.

\subsection{Software Attestation}
\label{subsection:attestation}

A research topic that exists in parallel to the detection and cooperation
duality is software attestation. Software attestation offers algorithms for
exchanging information about the software that is running on a node, and aims
to identify nodes that can no longer prove that their content is unaltered.
Examples of evolving algorithms to implement this functionality include SWATT
\cite{seshadri2004swatt}, ICE/SCUBA \cite{seshadri2006scuba} and SAKE
\cite{seshadri2008sake}.

Software attestation is hard and many details can cause an algorithm to fail.
Interesting is the discussion surrounding this topic that was started by
\cite{castelluccia2009difficulty}, in response to the previously mentioned
papers. In \cite{perrig2010refutation} the authors of the original papers
counter many of the objections made to their work, but the general feeling is
that even in the best conditions, there is always a way to circumvent even the
most ingenious algorithm to perform software-based attestation.

\section{Analysis of ID Algorithms}
\label{analysis}
